\newcommand{\GPH}{Geophysics}
\newcommand{\GJI}{Geophys. J. Int.}
\newcommand{\GJRAS}{Geophys. J. R. Astr. Soc.}
\newcommand{\GRL}{Geophys. Res. Let.}
\newcommand{\JGR}{J. Geophys. Res.}
\newcommand{\EPS}{Earth, Planets, Space}
\newcommand{\JGG}{Jour. Geomag. Geoelec.}
\newcommand{\IP}{Inv. Prob.}
\newcommand{\PEPI}{Phys. Earth Planet. Inter.}

\begin{thebibliography}{}

\bibitem{}
Avdev, D.~B., 2005.
Three-dimensional electromagnetic modelling and inversion from
theory to application
{\em Surv. Geophys.}, {\bf 26}, 767-799.
%
%\bibitem{}
%Bennett, A.F., Chua, B.S. and Leslie, L.M. 1996
%Generalized inversion of a global numerical weather prediction model
%{\em Meterology and Atmospheric Physics} {\bf 60} 165-178
%
\bibitem{}
Chua, B.~S., and Bennett, A.~F. 2001
An inverse ocean modeling system
{\em Ocean Modeling} {\bf 3} 137-165
%
\bibitem{}
Constable, S., Parker, R.~L., and Constable, C.~G. 1987
Occam's inversion: A practical algorithm for generating smooth models
from
electromagnetic sounding data
{\em \GPH} {\bf 52} 289-300
%
\bibitem{}
de Lugao, P., Portniaguine, O., and Zhdanov, M.~S. 1997
Fast and stable two-dimensional
inversion of magnetotelluric data
{\em \JGG} {\bf 49} 1437-1454
%
\bibitem{}
Egbert, G.~D., Bennett, A.~F., Foreman, M.~G.~G. 1994
Topex/Poseidon tides
estimated using a global inverse model
{\em \JGR} {\bf 99} 24821--24852
%
\bibitem{}
Haber, E. 2005
Quasi-Newton Methods for Large-Scale Electromagnetic Inverse Problems
{\em \IP} {\bf 21} 305-323
%
\bibitem{}
Kelbert et al., 2008
%
\bibitem{}
Mackie, R.~L and Madden, T.~R. 1993
Conjugate direction relaxation solutions for
3-D magnetotelluric modeling
{\em \GPH} {\bf 58} 1052-1057
%
\bibitem{}
Marquardt, D.~W. 1963
An Algorithm for least-squares estimation of nonlinear parameters
{\em J. Soc. Indust. Appl. Math} {\bf 11} 431-441
%
\bibitem{}
Newman, G.~A., and Alumbaugh, D.~L. 1997
Three-dimensional massively parallel
electromagnetic inversion-I. Theory
{\em \GJI} {\bf 128} 340-354
%
\bibitem{}
Newman, G.~A., and Alumbaugh, D.~L. 2000
Three-Dimensional Magnetotelluric Inversion
Using Non-Linear Conjugate Gradients
{\em \GJI} 140 410-424
%
\bibitem{}
Newman, G.~A., and Boggs, P.~T. 2004
Solution Accelerators for Large-Scale Three-
Dimensional Electromagnetic Inverse Problem
{\em \IP} 20 s151-s170
%
\bibitem{}
Nocedal, J., and Wright, S. 1999
{\em Numerical Optimization}
(New York: Springer-Verlag)
%
\bibitem{}
Press, W.~H., Flanerrry, B.~P., Teukolsky, S.~A., Vetterling, W.~T. 1986
{\em Numerical Recipes: The Art of Scientific Computing}
(Cambridge: Cambridge University Press)
%
\bibitem{}
Rodi, W.~L. 1976
A technique for improving the accuracy of finite element solutions for
magnetotelluric data
{\em \GJRAS} {\bf 44} 483-506
%
\bibitem{}
Rodi, W.~L., and Mackie, R.~L. 2001
Nonlinear Conjugate Gradients Algorithm for 2-D Magnetotelluric
Inversion
{\em \GPH} {\bf 66} 174-187
%
\bibitem{}
Parker, R.~L. 1994
{\it Geophysical Inverse Theory},
(Princeton NJ: Princeton University Press)
%
\bibitem{}
Siripunvaraporn, W. and Egbert, G. 2000
An Efficient Data-Subspace Inversion for Two-Dimensional
Magnetotelluric Data
{\em \GPH} {\bf 65} 791--803
%
\bibitem{}
Siripunvaraporn, W., Egbert, G., and Lenbury, Y. 2002
Numerical accuracy of magnetotelluric
modeling: A comparison of finite difference approximations
{\em \EPS} {\bf 54} 721-726
%
\bibitem{}
Siripunvaraporn, W., Uyeshima, M. and Egbert, G. 2004
Three-dimensional inversion for Network-Magnetotelluric data
{\em \EPS} {\bf 56} 893-902
%
\bibitem{}
Siripunvaraporn W. and Egbert, G. 2007
Data space conjugate  gradient inversion for 2-D Magnetotelluric  data
submitted to {\em GJI}
%
\bibitem{}
Siripunvaraporn, W., Egbert, G., Lenbury, Y., and Uyeshima, M. 2005
Three-Dimensional Magnetotelluric Inversion: Data Space Method
{\em \PEPI} {\bf 150} 3-14
%
\bibitem{}
Smith, J.~T. 1996
Conservative Modeling of 3-D Electromagnetic Fields, Part I, Properties
and Error Analysis
{\em Geophysics} {\bf 61} 1308-1318
%
\bibitem{}
Spitzer, K. 1998
The Three-dimensional sensitivity for surface and subsurface sources
{\em \GJI} {134} 736-746
%
\bibitem{}
Yee, K.~S. 1966
Numerical Solution of Initial Boundary Value Problems Involving
Maxwells Equations in Isotropic Media
{\em IEEE Trans. Ant. Prop.} {\bf AP-14} 302-307
%
\bibitem{}
Zhang, J., Mackie, R.~L., and Madden, T.~R. 1995
Three-dimensional resistivity forward modeling and inversion
using conjugate gradients
{\em Geophysics} {\bf 60} 1313-1325

\end{thebibliography}
